%%
%% starspan paper
%% Carlos A. Rueda-Velasquez
%% $Id$

\documentclass{elsart}
\usepackage{natbib,url,amssymb}

\journal{Computers \& Geosciences}


\begin{document}
\begin{frontmatter}

\title{STARSPAN: 
A Tool for Fast Selective Pixel Extraction 
from Remotely Sensed Data}
\thanks[label1]{Available from \url{http://starspan.casil.ucdavis.edu}}

\author{Carlos A. Rueda\corauthref{cor}},
\corauth[cor]{Corresponding author.}
\ead{carueda@ucdavis.edu}
\author{Jonathan Greenberg\thanksref{greenberg}}
\thanks[greenberg]{Present address: NASA Ames Research Center...}
\ead{greenberg@ucdavis.edu}

\address{Center for Spatial Technologies and Remote Sensing (CSTARS).
The Barn. University of California, Davis. One Shields Avenue, Davis, CA 95616}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
	
	Many research endeavours in environmental and ecological sciences [[?]]
	involve the utilization of a combined GIS and remote sensing approach. As
	the amount of data is steadily increasing due to the availability of more
	instruments as well as improved spatio-temporal-spectral resolutions, the
	risk for bottlenecks in data management and processing is also becoming a
	more important consideration that must be properly addressed for timely
	processing and results. Although there exist computer applications to help
	the scientist in these endeavours, there is an apparent absence of specific
	tools to integrate information from both GIS and remote sensing sources. In
	this paper we describe STARSPAN, a computer program focused on a specific
	yet crucial aspect of data integration which is the extraction of pixel data
	from spectral raster images according to specified geometry features.
	Development of the tool started as part of a project for mapping invasive
	plant species in the Sacramento-San Joaquin Delta Region (approximately
	$3400$~km$^2$ in California's Central Valley) using field data and airborne
	hyperspectral imagery. STARSPAN in implemented in C++ and developed under
	object-oriented design principles so it can be easily extended with new
	functionalities. As an open source project, it can be freely downloaded from
	\url{http://starspan.casil.ucdavis.edu}. In the context of real examples, we
	explain the usage of the tool, how it can be extended, and propose further
	directions for its improvement.
	
	
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{keyword}
GIS \sep remote sensing \sep image processing
% PACS codes here, in the form: \PACS code \sep code
\end{keyword}

\end{frontmatter}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}\label{intro}

	STARSPAN is an open source computer program developed at Center of
	Spatial Technologies and Remote Sensing (CSTARS) for fast and flexible
	extraction of pixel data from spectral raster images. The tool was developed
	as part of a project for mapping invasive plant species using hyperspectral
	imagery \citep{ustin04}.
	
	Main inputs for the program are one or multiple raster files and a vector
	file defining a set of geometry features. The basic operation performed by
	STARSPAN is the extraction of the spectral data from the raster files
	whose pixels are geometrically contained in the geometry features (points,
	lines, polygons) in the vector data. Extracted pixel data are stored in a
	number of different formats, including ENVI \citep{envi} (standard image and
	spectral library) and CSV (comma-separated values), a common text-based
	format that can be imported into standard applications for further analysis.
	Other functionalities include the generation of band specific statistics for
	specified features, extraction of features into individual rasters
	(mini-raster generation), update options, and generation of paired pixel
	band values for subsequent calculation of calibration coefficients.
	
	An important goal is that of software extensibility, hence STARSPAN is
	being developed under object-oriented design principles so it can be easily
	extended with new functionalities deriving from the core extraction process.

	[[...]]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{A use case}

	[[describe use in delta project...]]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Usage}

In this section we provide a detailed description of the main operations
performed by STARSPAN. 
Figure \ref{fig-usage} contains the usage message given by the program.
\begin{figure}[!ht]
\centering
\scriptsize{
\begin{verbatim}
USAGE:
  starspan <inputs/commands/options>...

   inputs:
      -raster <filenames>...
      -vector <filename>
      -speclib <filename>
      -update-csv <filename>
      -update-dbf <filename>

   commands:
      -raster_field <name>
      -raster_directory <directory>
      -csv <name>
      -envi <name>
      -envisl <name>
      -stats outfile.csv [avg|mode|stdev|min|max]...
      -calbase <link> <filename> [<stats>...]
      -report
      -dbf <name>
      -dump_geometries <filename>
      -mr <prefix>
      -jtstest <filename>

   options:
      -fields field1 field2 ... fieldn
      -pixprop <minimum-pixel-proportion>
      -noColRow
      -noXY
      -fid <FID>
      -skip_invalid_polys
      -progress [value]
      -RID_as_given
      -verbose
      -ppoly
      -in
      -srs <srs>
\end{verbatim}
}
\label{fig-usage}
\caption{Usage message}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Inputs and general options}

Raster data is given through the \verb|-raster| option:

	\verb|-raster| $R_1\ R_2\ \cdots\ R_n$

where $R_1, R_2, \ldots, R_n$ are raster datasets recognized by the GDAL library.

Vector data is given with the \verb|-vector| option:
	
	\verb|-vector| $V$

where $V$ is a vector datasource recognized by the OGR library.

There are some general options ...

	\verb|-fields| field$_1$ field$_2\ \cdots$ field$_n$
	
Only the specified fields will be transferred from vector file $V$
to the output. Use spaces to separate the field names.  Example:

   \verb|starspan -vector V -raster R -csv my.csv  -fields species foo|

will extract the fields with names `species' and `foo.' If no field from the
vector metadata is desired, just give the keyword \verb|-fields| with no
arguments. By default, all fields from vector will be extracted.

	\verb|-pixprop| value

Minimum proportion of pixel area in intersection so that the pixel is included.
A value in [0.0, 1.0] must be given.
Only used in intersections resulting in polygons. 
By default, the pixel proportion is 0.5. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Statistics}

Command \verb|-stats| is used to generate basic band-wise statistics for each
intersecting feature. Example:

	\verb|starspan -vector V -raster R -stats mystats.csv avg mode|

Creates \verb|mystats.csv|, a CSV (comma-separated-values) file containing
computed statistics for each FID. In the above example, codes \verb|avg| and
\verb|mode|
indicate to compute the average and mode statistics. Available statistics 
are shown in Table \ref{table-stats}.
\begin{table}[!ht]
\centering
\caption{Statistics}
\begin{tabular}{|l|l|}
\hline 
 code        & computes for each feature\\
\hline
\verb|avg  |  &   average of pixels in feature\\
\verb|mode |  &   Most frequent value\\
\verb|stdev|  &   sample standard deviation of pixels in feature\\
\verb|min  |  &   minimum value of pixels in feature\\
\verb|max  |  &   maximum value of pixels in feature\\
\hline
\end{tabular}
\label{table-stats}
\end{table}

The columns in the resulting \verb|myfile.csv| are:

	\begin{itemize}
		\item Feature ID 
		\item Names of attributes from the given input vector $V$ associated to FID. 
				If option \verb|-fields| is given, only the given fields will be extracted. 
		\item Band names. Each name takes the form $s$\_Band\#, where $s$
				is the code of the statistics, eg. \verb|stdev|, and \# is the band number 
				according to given rasters.  
	\end{itemize}

Subsequent lines contain corresponding values for each FID.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Software requirements}

	In order to build STARSPAN the following libraries are required.
	
	\begin{itemize}
		\item GEOS, Geometry Engine \citep{geos}.
				This library is required to support geometry operations.
				
		\item GDAL, Geospatial Data Abstraction Library \citep{gdal}.
	\end{itemize}



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion and further work}

	[[...]]
	
	Currently STARSPAN has been tested and used on Linux based operating
	environments. On the works is to make the tool publicly available and easy
	to install on different operating systems.


	For up-to-date documentation, including installation instructions and user
	manual, please visit: \url{http://starspan.casil.ucdavis.edu}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Acknowledments}

	[[three categories \verb|:-)| :
		contract funding the project;
		libraries(GDAL, GEOS...);
		and testers: Emma, Solomon, George, who else?...]]




	
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{elsart-harv}
\bibliography{starspan}

\end{document}


